---
title: "STP598sta: Spatiotemporal Analysis"
subtitle: "Homework 3"
author: "Name: Your name; NetID: Your ID"
date: "Due 11:59pm Friday October 28, 2022"
output:
  pdf_document: default
  # html_document: default
  # word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
options(width = 1000)
```

## Question 1

Compute the coherence (generalized correlation), $\frac{\mathrm{cov}(Y_\ell({\bf s}), Y_{\ell'}({\bf s}+{\bf h}))}{\sqrt{\mathrm{cov}(Y_\ell({\bf s}), Y_\ell({\bf s}+{\bf h})) \mathrm{cov}(Y_{\ell'}({\bf s}), Y_{\ell'}({\bf s}+{\bf h})) }}$:

(a) for the cross-covariance $\Sigma_{{\bf Y}({\bf s}),{\bf Y}({\bf s}')} = C({\bf s}-{\bf s}') = \sum_{j=1}^p \rho_j({\bf s}-{\bf s}') T_j$.

(b) for the cross-covariance $C({\bf s}-{\bf s}') = \sum_{u=1}^r \rho_u({\bf s}-{\bf s}') T^{(u)}$.



## Question 2

Let $Y({\bf s}) = (Y_1({\bf s}), Y_2({\bf s}))^T$ be a bivariate process with a stationary cross-covariance matrix function
$$C({\bf s}-{\bf s}') = \begin{pmatrix} c_{11}({\bf s}-{\bf s}') & c_{12}({\bf s}-{\bf s}')\\ c_{12}({\bf s}'-{\bf s}) & c_{22}({\bf s}-{\bf s}') \end{pmatrix}$$
and a set of covariates ${\bf x}({\bf s})$. Let ${\bf y}=({\bf y}_1^T, {\bf y}_2^T)^T$ be the $2n\times 1$ data vector, with ${\bf y}_1^T = (y_1({\bf s}_1), \cdots, y_1({\bf s}_n))^T$ and ${\bf y}_2^T = (y_2({\bf s}_1), \cdots, y_2({\bf s}_n))^T$.

(a) Show that the cokriging predictor has the form $$\mathrm{E}[Y_1({\bf s}_0)|{\bf y}] = {\bf x}^T({\bf s}_0)\boldsymbol\beta + \boldsymbol\gamma^T \Sigma^{-1}({\bf y} - X\boldsymbol\beta)$$, with appropriate definitions of $\boldsymbol\gamma$ and $\Sigma$.


(b) Show further that if ${\bf s}_k$ is a site where $y_l({\bf s}_k)$ is observed, then for $l=1,2$, $\mathrm{E}[Y_l({\bf s}_k)|{\bf y}]=y_l({\bf s}_k)$ if and only if $\tau^2_l=0$.


## Question 3

For a moving average process of the form
$$x_t = w_{t-1} + 2 w_t +  w_{t+1}$$
where $w_t$ are independent with zero means and variance $\sigma^2_w$, determine the autocovariance and autocorrelation functions as a function of lag $h=s-t$ and plot the ACF as a function of $h$.


## Question 4

In this problem, we explore the difference between a random walk and a trend stationary process.

(a) Generate four series that are random walk with drift, $x_t = \delta t + \sum_{i=1}^t w_j$, of length $n=100$ with $\delta=0.01$ and $\sigma_w=1$. Call the data $x_t$ for $t=1,\cdots, 100$. Fit the regression $x_t = \beta t + w_t$ using least squares. Plot the data, the true mean function (i.e. $\mu_t = 0.01t$) and the fitted line, $\hat x_t = \hat\beta t$, on the same graph. Hint: The following `R` code may be useful.

```{r, eval=F}
par(mfrow=c(2,2), mar=c(2.5,2.5,0,0)+.5, mgp=c(1.6,.6,0)) # set up 
for (i in 1:4){
  x = ts(cumsum(rnorm(100,.01,1)))        # data
  regx = lm(x~0+time(x), na.action=NULL)  # regression
  plot(x, ylab='Random Walk w Drift')     # plots
  abline(a=0, b=.01, col=2, lty=2)        # true mean (red - dashed)
  abline(regx, col=4)                     # fitted line (blue -  solid)
}
```

(b) Generate four series of length $n=100$ that are linear trend plus noise, say $y_t =0.01t +w_t$, where $t$ and $w_t$ are as in part (a). Fit the regression $y_t = \beta t +w_t$ using least squares. Plot the data, the true mean function (i.e. $\mu_t=0.01 t$) and the fitted line, $\hat y_t = \hat\beta t$, on the same graph.

(c) Comment (what did you learn from this assignment).
