---
title: "STP598sta: Spatiotemporal Analysis"
subtitle: "Point-referenced Models and Areal Models"
author: "Shiwei Lan"
date: Fall 2020
output:
  html_document:
    fontsize: 16pt
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
options(width = 1000)
```

## R Resources

We use `R` <http://cran.r-project.org/> for some demos in this lecture. `RStudio` <http://www.rstudio.com> is also recommended. Actually, this html demo is written RStudio using `RMarkdown`.

## Point-referenced Models

In lecture 1, we have learned point-referenced data models.
Let's look at some examples. More specifically, we will use `R` to show **EDA**, **variogram** and **kriging**.
To do that, we need R package called `geoR`. Let's install it.

```{r, eval=FALSE}
install.packages('geoR')
```

After we install it, we load it.

```{r}
library('geoR')
```

Ok, now we are ready for some exploratory data analysis (EDA).

### EDA

Let's first look at the scallops data, which give the locations and scallop catches in the Atlantic waters off the coasts of New Jersey and Long Island, New York.

```{r}
myscallops<-read.table('https://www.counterpointstat.com/uploads/1/1/9/3/119383887/myscallops.txt',header=T)
```

If the link fails, try to download it from the course website.
Now we plot the locations on the map.
```{r,eval=FALSE}
install.packages('maps')
```

```{r}
library(maps)
map('usa',xlim=c(-75.5,-71),ylim=c(38.2,41.5))
points(myscallops$long,myscallops$lat,pch=20,cex=0.75)
```

It is often helpful to create image plots and place contour lines on the plot.
These provide a visualization of the relized spatial surface. It is necessary to interpolate data to 'gap' areas.
For this purpose, we use the `interp.new` function in the R package `akima`.
Then contour lines can be added to the plot using the `contour` function.

```{r,eval=FALSE}
install.packages('akima')
```

```{r}
library(akima)
int.scp = interp(myscallops$long,myscallops$lat,myscallops$lgcatch,extrap=T)
image(int.scp,xlim=range(myscallops$long),ylim=range(myscallops$lat),xlab='Longitude',ylab='Latitude')
contour(int.scp,add=T)
```

Another useful tool is the 3d surface plot by `persp` function:

```{r}
persp(int.scp,theta = -30, phi = 30,xlim=range(myscallops$long),ylim=range(myscallops$lat),xlab='Longitude',ylab='Latitude',zlab='logcatch')
```

### variograms

Now we estimate variograms. We will use `variog` function from `geoR` package.
This function takes `geodata` object as in put, which can be created using `as.geodata` function.

```{r}
obj = cbind(myscallops$long,myscallops$lat,myscallops$lgcatch)
scallops.geo=as.geodata(obj,coords.col = 1:2,data.col = 3)
```

Now, a variogram object is created using `variog` function.
And the robust estimator can be obtained by using 'modulus' type in `variog` function.

```{r}
scallops.var=variog(scallops.geo,estimator.type='classical')
# scallops.var
scallops.var.robust=variog(scallops.geo,estimator.type='modulus')
```

We can plot them together
```{r, fig.height = 4, fig.width = 9}
par(mfrow=c(1,2))
plot(scallops.var)
title('semivariogram')
plot(scallops.var.robust)
title('robust semivariogram')
```

Followingly, we use `variofit` function from `geoR` to estimate the sill, the range, and the nugget parameters under a specific covariance model.
We can specify the covariance model by `cov.model` option.

```{r}
scallops.var.fit=variofit(scallops.var.robust, ini.cov.pars=c(2.0,.5),cov.model='exponential',fix.nugget = F,nugget = 1.0)
print(scallops.var.fit)
```

A Bayesian approach is also available through `likfit` function in `geoR`.
```{r}
scallops.lik.fit=likfit(scallops.geo, ini.cov.pars = c(2.0,.5),cov.model='exponential',trend='cte',fix.nugget = F,nugget = 1.0,nospatial=T,lik.method='ML')
print(scallops.lik.fit)
```
where `trend="cte"` means a spatial regression model with constant mean. The estimated result is `r scallops.lik.fit$beta`.

### kriging

There are two built-in functions in `geoR` for kriging: `krige.conv` is for convential kriging; `krige.bayes` performs Bayesian kriging.
Note, `krige.bayes` is limited to fitting a spatial regression model with constant mean, an exponential covariance model, a flat prior for the constant trend level, the reciprocal (Jeffrey's) prior for the sill, and a discrete uniform prior for the nugget.

```{r, results='hide'}
scallops.bayes = krige.bayes(scallops.geo,locations = 'no',borders = NULL, model = model.control(trend.d='cte',cov.model = 'exponential'),prior = prior.control(beta.prior = 'flat',sigmasq.prior = 'reciprocal',tausq.rel.prior ='uniform', tausq.rel.discrete = seq(0,1,.01) ))
```

We obtain samples for the mean, the sill, the range, and the nugget.

```{r, fig.height = 7, fig.width = 9}
out=scallops.bayes$posterior$sample
par(mfrow=c(2,2))
hist(out$beta,breaks=20,freq=F,main=NULL)
lines(density(out$beta),col='blue',lwd=2)
title('mean')
hist(out$sigmasq,breaks=20,freq=F,main=NULL)
lines(density(out$sigmasq),col='blue',lwd=2)
title('sill')
hist(out$phi,breaks=20,freq=F,main=NULL)
lines(density(out$phi),col='blue',lwd=2)
title('range')
hist(out$tausq,breaks=20,freq=F,main=NULL)
lines(density(out$tausq),col='blue',lwd=2)
title('nugget')
```

## Areal Models

In this section, we use some `R` function to obtain proximity matrix, compute Moran's and Geary's statistics and fit CAR and SAR models.
We will the `R` package named `spdep`. Let's first install and load it.

```{r, eval=F}
install.packages('spdep')
install.packages('maptools')
install.packages('RColorBrewer')
```

```{r}
library(spdep)
library(maptools)
library(classInt)
library(RColorBrewer)
```

First, let's compute the proximity (adjacency) matrix of US states.
```{r}
usa.state=map(database="state", fill=TRUE, plot=FALSE)
state.ID <- sapply(strsplit(usa.state$names, ":"), function(x) x[1])
usa.poly = map2SpatialPolygons(usa.state, IDs=state.ID)
usa.nb = poly2nb(usa.poly)
usa.adj.mat = nb2mat(usa.nb, style="B")
```

Constructing adjacency matrices for counties in a U.S. state is a little different because of the naming convention in `map`.
Now let's take Minnesota for example.
```{r}
mn.county = map("county", "minnesota", fill=TRUE, plot=FALSE)
county.ID <- sapply(strsplit(mn.county$names, ","), function(x) x[2])
mn.poly = map2SpatialPolygons(mn.county, IDs=county.ID)
mn.nb = poly2nb(mn.poly)
mn.adj.mat = nb2mat(mn.nb, style="B")
```
One can easily retrieve neighbors of a country. Let's obtain neighbors of `Winona` county in Minnesota.
```{r}
mn.region.id <- attr(mn.nb, "region.id")
winona.neighbors.index = mn.nb[[match("winona", mn.region.id)]]
winona.neighbors = rownames(mn.adj.mat[winona.neighbors.index,])
print(winona.neighbors)
```

### Moran's I and Geary's C

Now we can use `moran.test` and `geary.test` functions in the `spdep` package to obtain Moran's I and Geary's C statistics. Let's use the SAT scores data for illustration. First we need to convert `usa.nb` to a `listw` object in `R`.

```{r}
usa.listw = nb2listw(usa.nb, style="W")
state.sat.scores<-read.table('https://www.counterpointstat.com/uploads/1/1/9/3/119383887/state-sat.dat',header=F)
colnames(state.sat.scores) <- c('STATE','VERBAL','MATH','PERCT')
# remove alaska and hawaii
x = ((state.sat.scores$STATE=="alaska") | (state.sat.scores$STATE=="hawaii") | (state.sat.scores$STATE=="us"))
index = c(1:nrow(state.sat.scores))[x]
state.sat.scores.contig = state.sat.scores[-index,]
```

Next we use `moran.test` function to obtain
```{r}
moran.out=moran.test(state.sat.scores.contig$VERBAL, listw=usa.listw)
moran.I = moran.out$estimate[1]
moran.I.se = sqrt(moran.out$estimate[3])
moran_tab=data.frame(I=moran.I, se=moran.I.se)
table(moran_tab)
```
Geary's C can be computed analogously using the `geary.test` function
```{r}
geary.out=geary.test(state.sat.scores.contig$VERBAL, listw=usa.listw)
geary.C = geary.out$estimate[1]
geary.C.se = sqrt(geary.out$estimate[3])
geary_tab=data.frame(C=geary.C, se=geary.C.se)
table(geary_tab)
```

### CAR

### SAR
